---
title: "R: Data Manipulation"
subtitle: "dplyr"
author: "Murray Cadzow"
date: "2 February 2016"
output: ioslides_presentation
---

## Resources

 - Hadley's evolving book "Advanced R Programming"
 
  http://adv-r.had.co.nz
  
 - Hadley's guide to developing packages:
 
  http://r-pkgs.had.co.nz/
  
 - Tidy data
 
  https://github.com/hadley/tidy-data
  
  http://vita.had.co.nz/papers/tidy-data.pdf


## dplyr commands (verbs)

 - Verbs
   - filter: keep specific rows
   - select: choose columns by name
   - arrange: reorder rows
   - mutate: add variables
   - summarise: reduces variables ot values
   
 - Documentation:
 
```{r, eval=FALSE}
browseVignettes('dplyr')
```


## Flights Example

```{r}
## install.packages("dplyr")
## install.packages("nycflights13")
library(nycflights13)
library(dplyr)
```

---

```{r}
# Single table verbs
flights
```

---

```{r}
filter(flights, dest == "IAH")
```

---

```{r}
select(flights, starts_with("arr"))
```

---

```{r}
arrange(flights, desc(arr_delay))
```

---

```{r}
mutate(flights, speed = distance / air_time * 60)
```

---

```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```



## Nesting

```{r}
hourly_delay <- filter(
  summarise(
    group_by(
      filter(
        flights,
        !is.na(dep_delay)
      ),
      year, month, day
    ),
    delay = mean(dep_delay),
    n=n()
  ),
  n > 10
)
```

---

```{r}
hourly_delay
```


## The pipe operator: `%>%`

 - Read the pipe as "then": `x %>% f(y) -> f(x,y)`
 
```{r}
hourly_delay <- flights %>%
  filter(!is.na(dep_delay)) %>%
  group_by(year, month, day) %>%
  summarise(
    delay = mean(dep_delay),
    n = n()
  ) %>%
  filter(n > 10)
```

---

```{r}
hourly_delay
```


## Have a play


---

Combine today with tomorrow...

---

```{r, warning=FALSE}
library(ggplot2)
flights %>% 
  filter(carrier == "UA", origin == "EWR") %>% 
  ggplot(. , aes(x = air_time, y = distance)) + 
  geom_point(aes(colour = dest))
```

