---
title       : (Re)Introducing dplyr and ggvis
subtitle    : Or "why Hadley Wickham is awesome, part 72..."
author      : Mik Black
job         : Group Meeting, 12 August 2014
framework   : mikIO        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : mathjax            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## RStudio video: dplyr and ggvis

 - Presentations on dplyr (Hadley) and ggvis (Winston Chang) from July 30:
 
      Video: http://info.rstudio.net/n0N20NH02JCS0NX0X000Y74
      
      Code: https://github.com/rstudio/webinars

   - `dplyr` expands on (refines?) Hadley's previous packages for data manipulation (`reshape` and `plyr`).
 
   - ggvis provides the ability to embed interactive graphics into documents.

---

## Overview (stolen from talk)

<center><img src = 'PNG/had1.png' height=450></img></center>

### http://info.rstudio.net/n0N20NH02JCS0NX0X000Y74

---

## Resources

 - Hadley's evolving book "Advanced R Programming"
 
  http://adv-r.had.co.nz
  
 - Hadley's guide to developing packages:
 
  http://r-pkgs.had.co.nz/
  
 - Tidy data
 
  https://github.com/hadley/tidy-data
  
  http://vita.had.co.nz/papers/tidy-data.pdf

---

## dplyr

 - `dplyr` is all about 

---

## dplyr commands (verbs)

 - Verbs
   - filter: keep specific rows
   - select: choose columns by name
   - arrange: reorder rows
   - mutate: add variables
   - summarise: reduces variables ot values
   
 - Documentation:
 
```{r, eval=FALSE}
browseVignettes('dplyr')
```

---

## Example from webinar

```{r}
## Install the nycflights13 package via:
## install_github("hadley/nycflights13")
library(nycflights13)
library(dplyr)
```

---

```{r}
# Single table verbs
flights
```

---

```{r}
filter(flights, dest == "IAH")
```

---

```{r}
select(flights, starts_with("arr"))
```

---

```{r}
arrange(flights, desc(arr_delay))
```

---

```{r}
mutate(flights, speed = distance / air_time * 60)
```

---

```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

---

## Bad code - Mik style...

```{r}
hourly_delay <- filter(
  summarise(
    group_by(
      filter(
        flights,
        !is.na(dep_delay)
      ),
      year, month, day
    ),
    delay = mean(dep_delay),
    n=n()
  ),
  n > 10
)
```

---

```{r}
hourly_delay
```

---

## The pipe operator: `%>%`

 - Read the pipe as "then": `x %>% f(y) -> f(x,y)`
 
```{r}
hourly_delay <- flights %>%
  filter(!is.na(dep_delay)) %>%
  group_by(year, month, day) %>%
  summarise(
    delay = mean(dep_delay),
    n = n()
  ) %>%
  filter(n > 10)
```

---

```{r}
hourly_delay
```

---

## ggvis (example from webinar)

```{r} 
library(nycflights13)
library(dplyr)
library(ggvis)
library(lubridate)
library(knitr)

## Summarize and join daily flight and weather data
daily <- flights %>%
  filter(origin == "EWR") %>%
  group_by(year, month, day) %>%
  summarise(
    delay = mean(dep_delay, na.rm = TRUE),
    cancelled = mean(is.na(dep_delay))
  )
```

### https://github.com/rstudio/webinars/blob/master/2014-01/2-ggvis.R

---

## ggvis (example from webinar)

```{r}
daily_weather <- weather %>%
  filter(origin == "EWR") %>%
  group_by(year, month, day) %>%
  summarise(
    temp = mean(temp, na.rm = TRUE),
    wind = mean(wind_speed, na.rm = TRUE),
    precip = sum(precip, na.rm = TRUE)
  )
both <- daily %>%
  inner_join(daily_weather) %>%
  ungroup() %>%
  mutate(date = as.Date(ISOdate(year, month, day)))
```

### https://github.com/rstudio/webinars/blob/master/2014-01/2-ggvis.R

---

## Pipes in ggvis

```{r, eval=FALSE}
## NB - won't embed into Slidify presentation...
both %>%
  ggvis(x = ~temp, y = ~delay) %>%
  layer_points() %>%
  layer_smooths()
```
