---
title: "Packages - Lesson 2"
author: "Murray Cadzow"
date: "29 January 2016"
output: ioslides_presentation
---

## Attribution

Much of this content is adapted from Hadley Wickham's R packages book
http://r-pkgs.had.co.nz/tests.html

and from Karl Broman
http://kbroman.org/pkg_primer/pages/tests.html

## Getting set up

This will carry on with the example from lesson 1

```{r eval = FALSE}
install.packages('testthat')

library(testthat)

```

change your working dir to 'cats'

```{r, eval = FALSE}
setwd("cats")
```

## Initialise the test directory

```{r, eval = FALSE}

devtools::use_testthat()

```

This creates

* A tests/testthat directory

* Adds testthat to the Suggests field in the DESCRIPTION

* Creates a file tests/testthat.R that runs all your tests are when R CMD check runs

## The Tests Dir

```{r, eval = FALSE}
list.files(path = "tests/", recursive=TRUE, include.dirs=TRUE)
```

* The tests dir has tests/testthat.R and tests/testthat/

* We create our tests to be run in tests/testthat/

* The filenames start with test, eg test_cat_function.R

## tests/testthat.R

This file tells test that to load the testthat library when testing

and to go through the tests that are stored in tests/testthat/

## Test structure

Test files contain 

* __context__ - a description of what is being tested

* one or more tests

Tests are organised hierarchically: 

__expectations__ are grouped into __tests__ which are organised in __files__

```{r, eval=FALSE}
library(stringr)
context("String length")

test_that("str_length is number of characters", {
  expect_equal(str_length("a"), 1)
  expect_equal(str_length("ab"), 2)
  expect_equal(str_length("abc"), 3)
})
```

## Tests

A test groups together multiple expectations to test output

This can range from a simple function through to a complex multi-argument function

Sometimes called a __unit__ test because it tests one unit of functionality


## Expectations

The 'atom' of testing

Describes the output of a computation

Does it produce what it should?

Does it produce an error code?

Expectations are functions that start with expect_

## Expectations cont.

Expectations make a binary assertion about what a function does

Structure:

  * start with __expect\___

  * have two arguments, 1st: actual result, 2nd what was expected
  
  * if actual and expected don't match testthat throws an error

## Expectations cont. 2

the common expectations that we can use are:

```{r, eval = FALSE}
expect_equal()

expect_identical()

expect_match()

expect_output()

expect_message()

expect_error()
```

There are many others too

##


```{r, eval = FALSE}
expect_equal(10, 10)

expect_equal(10, 10 + 1e-7)

expect_equal(10, 11)
#> Error: 10 not equal to 11

```

expect_equal uses all.equal() which allows a numerical tolerance 

For exact comparisons use expect_identical(), which uses identical() 

```{r eval = FALSE}
expect_equal(10, 10 + 1e-7)

expect_identical(10, 10 + 1e-7)
#> Error: 10 is not identical to 10 + 1e-07. Differences: 
#> Objects equal but not identical
```

##

```{r, eval = FALSE}
string <- "Testing is fun!"

expect_match(string, "Testing") 

# Fails, match is case-sensitive
expect_match(string, "testing")
#> Error: string does not match 'testing'. Actual value: "Testing is fun!"

# Additional arguments are passed to grepl:
expect_match(string, "testing", ignore.case = TRUE)
```

##

Variations of expect_match() allow you to check for:

```{r, eval=FALSE}
# checks printed output
expect_output()

# checks messages
expect_message()

#checks warnings
expect_warning()

# checks errors
expect_error()
```

## What to test

If you find a bug, write a test __then__ fix

How can someone break your function?
  
  * What happens if an argument isn't passed?
  
  * What happens if someone passes the wrong data structure/type?
  
  * What happens with NAs?
  
  * Do your conditionals work as expected?
  
What situations does your function need to handle and what can you let fail?

## Cats example

Lets make a test for our cats function