---
title: "R-based web apps with Shiny"
author: "Mik Black"
date: "4 February 2016"
output: html_document
---

## Rstudio links

http://shiny.rstudio.com/

http://shiny.rstudio.com/tutorial/

## Load shiny package

```{r}
library(shiny)
```

## Example ui.R code

Taken from: http://shiny.rstudio.com/tutorial/

```{r, eval=FALSE}


shinyUI(bootstrapPage(

  selectInput(inputId = "n_breaks",
      label = "Number of bins in histogram (approximate):",
      choices = c(10, 20, 35, 50),
      selected = 20),

  checkboxInput(inputId = "individual_obs",
      label = strong("Show individual observations"),
      value = FALSE),

  checkboxInput(inputId = "density",
      label = strong("Show density estimate"),
      value = FALSE),

  plotOutput(outputId = "main_plot", height = "300px"),

  # Display this only if the density is shown
  conditionalPanel(condition = "input.density == true",
    sliderInput(inputId = "bw_adjust",
        label = "Bandwidth adjustment:",
        min = 0.2, max = 2, value = 1, step = 0.2)
  )

))

```

## Example server.R code

Taken from: http://shiny.rstudio.com/tutorial/

```{r, eval=FALSE}
shinyServer(function(input, output) {

  output$main_plot <- renderPlot({

    hist(faithful$eruptions,
      probability = TRUE,
      breaks = as.numeric(input$n_breaks),
      xlab = "Duration (minutes)",
      main = "Geyser eruption duration")

    if (input$individual_obs) {
      rug(faithful$eruptions)
    }

    if (input$density) {
      dens <- density(faithful$eruptions,
          adjust = input$bw_adjust)
      lines(dens, col = "blue")
    }

  })
})
```

## Can we make it more simple?

Start with the "cars" data set from base R:

```{r}
data(cars)
dim(cars)
head(cars)
```

## What might we like to plot?

```{r, fig.width=10}
par(mfrow=c(1,2))
hist(cars$speed)
hist(cars$dist)
```

## Different syntax

Remember that we can also specify columns via: `hist(cars[,"speed"])` which will come in handy later:

```{r}
hist(cars[ , "speed"])
```

## Creating a VERY basic shiny app: content

server.R file:

```{r, eval=FALSE}
## Load the "cars" data set
data(cars)

## Create the function to be excuted by shiny 
shinyServer(function(input, output) {
  
## Define an object called "main_plot" that will be drawn by the "renderPlot" function
  output$main_plot <- renderPlot({

## Specify the plot that will be generated
      hist(cars[,"speed"])
  })
  
})
```

## Creating a VERY basic shiny app: interface

Initially we have no need for an interface - the plot we are generating is fixed.

ui.R file:

```{r, eval=FALSE}
shinyUI(bootstrapPage(
  
## Tell shiny what to plot - the "main_plot" object associated with "output" generated
## by the server.R file
  plotOutput(outputId = "main_plot")
  
))

```

## Running the app

Create a subdirectory called "carsApp", and place the ui.R and server.R files in that directory.

In R, load the shiny library, and run the app:

```{r, eval=FALSE}
library(shiny)
runApp('carsApp')
```

If you are working in RStudio, the `runApp()` command will open your app in a built-in web browser, otherwise it will open in your usual system browser (e.g., Firefox).

## That app is boring!!  (and ugly)

All the `runApp()` command did was create an ugly plot in a web browser.

We won't fix the "ugly" just yet, but lets improve the interactivity.

## Adding interactivity: ui.R

Add an option to choose which variable to plot.

First update the interface:

```{r, eval=FALSE}
shinyUI(bootstrapPage(
  
  selectInput(inputId = "plot_var",
              label = "Variable to plot",
              choices = c("speed", "dist"),
              selected = "speed"),
    
## Tell shiny what to plot - the "main_plot" object associated with "output" generated
## by the server.R file
  plotOutput(outputId = "main_plot")
  
))
```

Then modify the server.R file to accommodate the new information. Specifically, we edit the line:

```{r, eval=FALSE}
hist(cars[,"speed"])
```

and replace it with 

```{r, eval=FALSE}
hist(cars[,input$plot_var])
```

This tells the `hist` command to plot the data specified by `plot_var` from the `input` object (specified in ui.R).

```{r, eval=FALSE}
## Load the "cars" data set
data(cars)

## Create the function to be excuted by shiny 
shinyServer(function(input, output) {
  
## Define an object called "main_plot" that will be drawn by the "renderPlot" function
  output$main_plot <- renderPlot({

## Specify the plot that will be generated
      hist(cars[,input$plot_var])
  })
  
})
```

## Test the new app

```{r, eval=FALSE}
runApp('carsApp')
```

Now we've got some interactivity!  Time work on the "ugly"...

## Change the page style

ui.R file:

```{r, eval=FALSE}
shinyUI(pageWithSidebar(
  
## Give the application a title
  headerPanel("My awesome plotting app"),
  
## Put the input controls in a sidebar 
  sidebarPanel(
    
  selectInput(inputId = "plot_var",
              label = "Variable to plot",
              choices = c("speed", "dist"),
              selected = "speed")
  ),

## Put the outout in the main panel
  mainPanel(

    ## Give the panel a name
    h4("Plot of cars data"),
    
    ## Tell shiny what to plot - the "main_plot" object associated with "output" generated
    ## by the server.R file
    plotOutput(outputId = "main_plot")
  )
))

## Lets get some more interactivity

Add the following to the ui.R file below the current `selectInput` statement:

```{r, eval=FALSE}
sliderInput(inputId = "nbins",
            label = "Number of bins",
            min=2,
            max=15,
            value=10,
            round=TRUE)
```

Note that you'll need to add a comma to the last line of the `selectInput`, that is:

```{r, eval=FALSE}
selected = "speed"),
```

Now we need to update the server.R file to use this new information.  Replace the line:

```{r, eval=FALSE}
hist(cars[,input$plot_var])
```

with

```{r, eval=FALSE}
hist(cars[,input$plot_var], breaks = input$nbins)
```

Run the app, and see what happens!

```{r, eval=FALSE}
runApp('carsApp')
```

## Getting fancier - adding tabs:

Create something new to plot in server.R (add below the existing plotting code):

```{r, eval=FALSE}
## Define a second plot
  output$second_plot <- renderPlot({

## Specify the plot that will be generated
      boxplot(cars[,input$plot_var])
  })
```

Now alter the "mainPanel" section of ui.R to add information about tabs:

```{r, eval=FALSE}
  mainPanel(

    ## Add tabs
    tabsetPanel(

      ## Define a tab for the histogram
      tabPanel(
        h4("Plot of cars data"),
        plotOutput(outputId = "main_plot")
      ),

     ## And a tab for the boxplot
      tabPanel(
        h4("Boxplot of cars data"),
        plotOutput(outputId = "second_plot")
      )
     
    )
  )
```

Note that the `outputId` defined in `server.R` is being used to determine which output to include on each tab.

Run the app - this is getting cooler.

```{r, eval=FALSE}
runApp('carsApp')
```

## Conditioning

Shiny also provides the ability to "conditionally" display input selectors.

In our current app, the slider input remains visible when the boxplot tab is being viewed, even though it is not relevant for that plot (and moved it has no impact).

To ensure that the slider only appears for the histogram panel, alter the ui.R file to specify when it should be displayed.

The sidebarPanel section of ui.r now becomes:

```{r, eval=FALSE}
sidebarPanel(
    
  selectInput(inputId = "plot_var",
              label = "Variable to plot",
              choices = c("speed", "dist"),
              selected = "speed"),

  ## Add a conditionalPanel containing the slider - define it as "condition 1"
  conditionalPanel(condition="input.conditionedPanels==1",
                   sliderInput(inputId = "nbins",
                               label = "Number of bins",
                               min=2,
                               max=15,
                               value=10,
                               round=TRUE)
                  )
  ),
```

In the mainPanel section, specify when "condition 1" occurs:

```{r, eval=FALSE}
  mainPanel(

    tabsetPanel(

      tabPanel(
        h4("Plot of cars data"),
        plotOutput(outputId = "main_plot")
        ## Define the value of this tab
        , value=1
      ),

      tabPanel(
        h4("Boxplot of cars data"),
        plotOutput(outputId = "second_plot")
        ## Define the value of this tab (not actualled necessary - next line not needed)
        , value=0
      )
      ## Name the panel to correspond to the codition defined in sidebarPanel
      , id = "conditionedPanels"
    )
  )
```

Now when you run the app, the slider only appears in the sidebarPanel when the histrogram tab is being displayed.  Much tidier....











